---
title: "Human Activity Recognition"
author: "Bart Lenoir"
date: "22 November 2015"
output: 
  html_document: 
    keep_md: yes
---

### Synopsis

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

### Libraries

```{r}
library(caret)
library(randomForest)
```

### Data Source

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har

```{r}
setwd("/Users/bartlenoir/gitrepo/HumanActivityRecognition")

# download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-training.csv")
pmlTraining <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!", ""))

# download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-testing.csv")
pmlTesting <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!", ""))
```

**View data.**

```{r}
dim(pmlTraining)
str(pmlTraining, list.len=10)
# colnames(pmlTraining)
```

```{r}
dim(pmlTesting)
str(pmlTesting, list.len=10)
# colnames(pmlTesting)
```

### Data Processing

**Check for complete cases.**

```{r}
sum(complete.cases(pmlTraining))
sum(complete.cases(pmlTesting))
```

**Remove incomplete columns.**

```{r}
# removing columns true zero variance
pmlTraining <- pmlTraining[, colSums(is.na(pmlTraining)) == 0]
pmlTesting <- pmlTesting[, colSums(is.na(pmlTesting)) == 0]
```

**Recheck for complete cases.**

```{r}
sum(complete.cases(pmlTraining))
sum(complete.cases(pmlTesting))
```

**Remove irrelevant columns.**

```{r}
pmlTraining <- pmlTraining[,-c(1:7)]
pmlTesting <- pmlTesting[,-c(1:7)]
```

**Review data.**

```{r}
# check variance
sum(nearZeroVar(pmlTraining, saveMetrics = TRUE)$zeroVar == T)
dim(pmlTraining)
dim(pmlTesting)
head(pmlTraining, n = 2)
head(pmlTesting, n = 2)
```

**Prepare for cross validation (random subsampling without replacement).**

```{r}
trainingDataSubSet <- createDataPartition(y=pmlTraining$classe, p=0.70, list=FALSE)
pmlTrainingSubSet <- pmlTraining[trainingDataSubSet, ] 
pmlTestingSubSet <- pmlTraining[-trainingDataSubSet, ]
```

```{r}
dim(pmlTrainingSubSet)
dim(pmlTestingSubSet)
```

### Results

**Create model.**

```{r}
set.seed(123456789)

modFit <- randomForest(classe ~ ., data = pmlTrainingSubSet, method="class")

print(modFit)

plot(modFit)
```

**Plot margin of observations (positive margin means correct classification).**

```{r}
plot(margin(modFit,pmlTrainingSubSet$classe))
```

**Run model on 'pmlTestingSubSet'.**

```{r}
# fine tuning model doesn't improve accuracy
# tuneRF(subset(pmlTrainingSubSet, select = -c(classe)), pmlTrainingSubSet$classe, doBest = TRUE)
# varImpPlot(modFit, pch = 19)

pmlPred <- predict(modFit, newdata = pmlTestingSubSet, type = "class")
```

**Expected out-of-sample error:**

```{r}
confusionMatrix(pmlPred, pmlTestingSubSet$classe)
```

**Run model on final 'pmlTesting'.**

```{r}
pmlPredFinal <- predict(modFit, newdata = pmlTesting, type="class")
pmlPredFinal
```

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(pmlPredFinal)
```